\pdftitle

### 2017-05-15 maxNt.tex
more discussion

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-15T17:29:54+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

Suppose our desired constraints are $\expe{\yxs } = c_1$ and
$\expe{\yxxs} = c_2$, the expectations being given by the fourth term of
\eqn~\eqref{eq:expe_products}. Applying \me\ to the distribution of the
average of the full population we find
\begin{equation}
  \label{eq:full_pop_av_example_ME}
  \p(\yXf = \yRf \| \yHa)  = \yL\,
\binom{N}{N \yRf}\exp\biggl[
\yL_2\, \frac{\yRf\,(N \yRf-1)}{N-1} + \yL_1\, \yRf\biggr]
\end{equation}
with such parameters $\yL$, $\yL_1$, $\yL_2$ as to satisfy normalization
and constraints. We call this state of knowledge $\yHa$. Marginalization by
\eqn~\eqref{eq:marginal} then gives
\begin{equation}
  \label{eq:full_pop_av_example_ME_marg}
  \p(\yxs = \yrs \| \yHa)  = \yL
  \sum_{N\yRf=0}^N
  \binom{n}{n\yrs} \binom{N-n}{N\yRf-n\yrs} \exp\biggl[
\yL_2\, \frac{\yRf\,(N \yRf-1)}{N-1} + \yL_1\, \yRf\biggr].
\end{equation}
Applying \me\ to the distribution of the average of the sample, using the
third term of \eqn~\eqref{eq:expe_products}, we find
\begin{equation}
  \label{eq:sample_av_example_ME}
  \p(\yxs = \yrs \| \yHb)  = \yl\,
  \binom{n}{n\yrs}
  \exp\biggl[\yl_2\, \frac{\yrs\,(n \yrs-1)}{n-1} + \yl_1\, \yrs\biggr]
\end{equation}
%\setlength{\intextsep}{0.5ex}%
%\begin{wrapfigure}{r}{0.5\linewidth}
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\linewidth]{different_maxent_pop_sample_100.pdf}%
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.45$,
  $\expe{\yxxs \|\yH}=0.35$, and $N=5\,000$, $n=100$.}
\label{fig:diff_maxent_pop_sample}
\end{figure}%maxent_pop_or_sample.nb
%\end{wrapfigure}
with such parameters $\yl$, $\yl_1$, $\yl_2$ as to satisfy the constraints.
We call this state of knowledge $\yHb$.

The resulting distributions $\p(\yxs = \yrs \| \yHa)$,
$\p(\yxs = \yrs \| \yHb)$ are different.
Figure~\ref{fig:diff_maxent_pop_sample} shows an example, for values
$N=5\,000$, $n=100$, $c_1=0.45$, $c_2=0.35$, yielding $\yL_1 = -13\,321.9$,
$\yL_2 = 13\,321.5$, $\yl_1=-269.3$, $\yl_2 = 268.9$, and appropriate
normalization constants.

@@

Which application of the \me\ method should we choose, then?

@@

The quandary we found in the present note has at least two bright sides.
The first is that the two different \me\ applications may give numerically
similar results in some cases. One could thus be used as an approximation
of the other, depending on the circumstances -- \eg\ if probability tails
are unimportant. See \fig~\ref{fig:diff_maxent_pop_sample_realdata} for an
example.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5\,000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
The second is that the possibility of using two different distributions is
not a physical contradiction. Similar situations arise in statistical
mechanics. It is known that if a system is described by a \me\ Gibbs state,
its subsystems need not be \cite{maesetal1999}. A quandary somehow similar
to ours also appears in the statistical description of an equilibrium state
at the end of a non-equilibrium process: we can describe our knowledge
about it either by a Gibbs distribution, or by the Liouville-evolved Gibbs
distribution associated with equilibrium state at the beginning of the
process. The two descriptions differ -- even though the final physical
state is exactly the same \cite[\sect~4]{jaynes1985d_r1993}. The appearance
of this difference is natural: in one case we can make sharper predictions
about the state thanks to our knowledge of its preceding dynamics.
\mynote{but in this case both distributions are immensely sharp, and this
  doesn't affect predictions}

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-15T17:53:53+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

Suppose our desired constraints are $\expe{\yxs } = c_1$ and
$\expe{\yxxs} = c_2$, the expectations being given by the fourth term of
\eqn~\eqref{eq:expe_products}. Applying \me\ to the distribution of the
average of the full population we find
\begin{equation}
  \label{eq:full_pop_av_example_ME}
  \p(\yXf = \yRf \| \yHa)  = \yL\,
\binom{N}{N \yRf}\exp\biggl[
\yL_2\, \frac{\yRf\,(N \yRf-1)}{N-1} + \yL_1\, \yRf\biggr]
\end{equation}
with such parameters $\yL$, $\yL_1$, $\yL_2$ as to satisfy normalization
and constraints. We call this state of knowledge $\yHa$. Marginalization by
\eqn~\eqref{eq:marginal} then gives
\begin{equation}
  \label{eq:full_pop_av_example_ME_marg}
  \p(\yxs = \yrs \| \yHa)  = \yL
  \sum_{N\yRf=0}^N
  \binom{n}{n\yrs} \binom{N-n}{N\yRf-n\yrs} \exp\biggl[
\yL_2\, \frac{\yRf\,(N \yRf-1)}{N-1} + \yL_1\, \yRf\biggr].
\end{equation}
Applying \me\ to the distribution of the average of the sample, using the
third term of \eqn~\eqref{eq:expe_products}, we find
\begin{equation}
  \label{eq:sample_av_example_ME}
  \p(\yxs = \yrs \| \yHb)  = \yl\,
  \binom{n}{n\yrs}
  \exp\biggl[\yl_2\, \frac{\yrs\,(n \yrs-1)}{n-1} + \yl_1\, \yrs\biggr]
\end{equation}
with such parameters $\yl$, $\yl_1$, $\yl_2$ as to satisfy the constraints.
We call this state of knowledge $\yHb$.
%\setlength{\intextsep}{0.5ex}%
%\begin{wrapfigure}{r}{0.5\linewidth}
\begin{figure}[!t]
\centering
\includegraphics[width=0.99\linewidth]{different_maxent_pop_sample_100.pdf}%
\caption{\mynote{shorten }Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.45$,
  $\expe{\yxxs \|\yH}=0.35$, and $N=5\,000$, $n=100$.}
\label{fig:diff_maxent_pop_sample}
\end{figure}%maxent_pop_or_sample.nb
% \end{wrapfigure}
\begin{figure}[!t]
\centering
\includegraphics[width=0.99\linewidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{\mynote{shorten }Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5\,000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb

The resulting distributions $\p(\yxs = \yrs \| \yHa)$,
$\p(\yxs = \yrs \| \yHb)$ are different.

Figure~\ref{fig:diff_maxent_pop_sample} shows one example, for values
$N=5\,000$, $n=100$, $c_1=0.45$, $c_2=0.35$, yielding $\yL_1 = -13\,321.9$,
$\yL_2 = 13\,321.5$, $\yl_1=-269.3$, $\yl_2 = 268.9$, and appropriate
normalization constants. Another example is shown in
\fig~\ref{fig:diff_maxent_pop_sample_realdata}, for neurobiologically more
realistic constraints $c_1=0.045$, $c_2=0.0025$, and $N=5\,000$, $n=200$,
yielding $\yL_1 = -16\,834.0$, $\yL_2 = 16\,825.7$, $\yl_1=-673.5$,
$\yl_2 = 665.1$. The discrepancy, although not as large as in the previous
example, is still relevant for the tails of the distributions.

@@

Where should we apply the \me\ method then, on the sample or on the
population?

@@

\iffalse
The quandary we found in the present note has at least two bright sides.
The first is that the two different \me\ applications may give numerically
similar results in some cases. One could thus be used as an approximation
of the other, depending on the circumstances -- \eg\ if probability tails
are unimportant. See \fig~\ref{fig:diff_maxent_pop_sample_realdata} for an
example.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5\,000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
The second is that
\fi
The possibility of using two different distributions is
not a physical contradiction. Similar situations arise in statistical
mechanics. It is known that if a system is described by a \me\ Gibbs state,
its subsystems need not be \cite{maesetal1999}. A quandary somehow similar
to ours also appears in the statistical description of an equilibrium state
at the end of a non-equilibrium process: we can describe our knowledge
about it either by a Gibbs distribution, or by the Liouville-evolved Gibbs
distribution associated with equilibrium state at the beginning of the
process. The two descriptions differ -- even though the final physical
state is exactly the same \cite[\sect~4]{jaynes1985d_r1993}. The appearance
of this difference is natural: in one case we can make sharper predictions
about the state thanks to our knowledge of its preceding dynamics.
\mynote{but in this case both distributions are immensely sharp, and this
  doesn't affect predictions}
